---
title: "Baseball Project Models"
author: "Joel Kopp"
date: "July 25, 2018"
output:
  html_document: 
    toc: yes
  pdf_document: default
urlcolor: cyan
editor_options: 
  chunk_output_type: console
---

***

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
options(scipen = 1, digits = 4, width = 80, fig.align = "center")
```

#Data Aggregation and Preparation

```{r message=FALSE, warning=FALSE}
library(dplyr)
```

Read in the datasets compiled by Sam Lahman.

```{r}
salaries = read.csv("Salaries.csv")
teams = read.csv("Teams.csv")
batting = read.csv("Batting.csv")
```

Reduce the dataset to only include data from the years 2000 to 2016.

```{r}
#Reduce data to Year >=2000 thru 2016
teams = teams[teams$yearID >= 2000 & teams$yearID < 2017,]
teams = droplevels(teams)
salaries = salaries[salaries$yearID >= 2000 & salaries$yearID < 2017,]
salaries = droplevels(salaries)
batting = batting[batting$yearID >= 2000 & batting$yearID < 2017,]
batting = droplevels(batting)
```

Verify the factor levels for `teamID` to ensure a seamless join.

```{r}
#Check identity levels
identical(levels(salaries$teamID), levels(teams$teamID))
identical(levels(batting$teamID), levels(teams$teamID))
```

Aggregate the player salaries by team to get team payroll per year.

```{r warning=FALSE}
# sum salaries by year by team
team_salaries = salaries %>% group_by(yearID, teamID) %>% summarize(salary = sum(as.numeric(salary)))
```

Aggregate player `RBI`, `GIDP`, and `IBB` by team per year.

```{r}
# sum batting by year by team
team_batting = batting %>% group_by(yearID, teamID) %>% 
                  summarise_at( 
                    .vars = vars(RBI, GIDP, IBB),
                    .funs = sum)
```

Join the aggregated salary and individual player batting dataframes to the main `teams` dataframe.

```{r message=FALSE}
# Now join it together
teams = inner_join(teams, team_salaries)
teams = inner_join(teams, team_batting)
```

***

#Advanced Sabermetric Data Calculations

```{r}
# Calculate team total bases
# Since a doubles, triples, and home runs are already counted as hits,
# multiply the number of bases in the extra-base hit by (no. bases - 1)
teams$TB = teams$H + teams$X2B + 2 * teams$X3B + 3 * teams$HR

# Calculate team slugging percentage
teams$SLG = teams$TB / teams$AB

# Calculate team on-base percentage
teams$OBP = (teams$H + teams$BB + teams$HBP) / (teams$AB + teams$BB + teams$HBP + teams$SF)

# Calculate OPS
teams$OPS = teams$SLG + teams$OBP

# Calculate team WHIP
teams$WHIP = (teams$BBA + teams$HA) / (teams$IPouts / 3)

# Calculate team batting average on balls in play
teams$BABIP = (teams$H - teams$HR) / (teams$AB - teams$SO - teams$HR + teams$SF)

# Calculate runs created
teams$RC = teams$TB * (teams$H + teams$BB) / (teams$AB + teams$BB)

# Calucate team singles
teams$X1B = teams$H - teams$X2B - teams$X3B - teams$HR

# Calculate team unintentional walks
teams$uBB = teams$BB - teams$IBB

# Calculate team wOBA
teams$wOBA = (0.69 * teams$uBB + 0.72 * teams$HBP + 0.89 * teams$X1B + 
                1.27 * teams$X2B + 1.62 * teams$X3B + 2.10 * teams$HR) / 
  (teams$AB + teams$BB - teams$IBB + teams$SF + teams$HBP)
```

---

#Potential Models

###Tests

```{r message=FALSE, warning=FALSE}
library(lmtest)

get_bp_decision = function(model, alpha) {
  decide = unname(bptest(model)$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_sw_decision = function(model, alpha) {
  decide = unname(shapiro.test(resid(model))$p.value < alpha)
  ifelse(decide, "Reject", "Fail to Reject")
}

get_num_params = function(model) {
  length(coef(model))
}

get_loocv_rmse = function(model) {
  sqrt(mean((resid(model) / (1 - hatvalues(model))) ^ 2))
}

get_adj_r2 = function(model) {
  summary(model)$adj.r.squared
}
```

```{r}
plot_fitted_resid = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  plot(fitted(model), resid(model), 
       col = pointcol, pch = 20, cex = 1.5,
       xlab = "Fitted", ylab = "Residuals", main = "Fitted vs Residuals")
  abline(h = 0, col = linecol, lwd = 2)
}

plot_qq = function(model, pointcol = "dodgerblue", linecol = "darkorange") {
  qqnorm(resid(model), col = pointcol, pch = 20, cex = 1.5)
  qqline(resid(model), col = linecol, lwd = 2)
}

test_all = function(model) {
  list(bp = get_bp_decision(model, 0.01),
  sw = get_sw_decision(model, 0.01),
  loocv = get_loocv_rmse(model),
  adjr2 = get_adj_r2(model),
  n = get_num_params(model))
}
```

###Model 1

```{r}
wins_add_full_model = lm(W ~ R + H + X2B + X3B + HR + BB + SO + SB + CS + HBP + SF + RA + ER + ERA + CG + SHO + SV + HA + HRA + BBA + SOA + E + DP + FP + salary + RBI + GIDP + IBB + TB + SLG + OBP + OPS + WHIP + BABIP + RC + X1B + uBB + wOBA, data = teams)

wins_sel_mod1 = step(wins_add_full_model, trace = 0)

test_all(wins_sel_mod1)
car::vif(wins_sel_mod1)
summary(wins_sel_mod1)

wins_sel_mod2 = lm(W ~ R + H + X2B + X3B + HR + CS + SF + RA + CG + SHO + SV + HA + HRA + BBA + E + FP + salary + GIDP + WHIP, data = teams)

test_all(wins_sel_mod2)
car::vif(wins_sel_mod2)
summary(wins_sel_mod2)

wins_sel_mod3 = lm(W ~ R + H + X2B + X3B + HR + CS + SF + RA + CG + SHO + SV + HA + HRA + E + FP + salary + GIDP + WHIP, data = teams)

test_all(wins_sel_mod3)
car::vif(wins_sel_mod3)
summary(wins_sel_mod3)

wins_sel_mod4 = lm(W ~ R + X3B + CS + RA + CG + SHO + SV + HRA + salary, data = teams)

(model_1_tests = test_all(wins_sel_mod4))
(model_1_vif = car::vif(wins_sel_mod4))
summary(wins_sel_mod4)
```

```{r fig.width = 10}
par(mfrow = c(1, 2))
plot_fitted_resid(wins_sel_mod4)
plot_qq(wins_sel_mod4)
```

Model 1 is a purely additive model and has an adjusted $R^2$ of `r model_1_tests$adjr2`, a LOOCV RMSE of `r model_1_tests$loocv`, with constant variance and normality confirmed at $\alpha = 0.01$.  The model was derived using an initial full additive model and reducing it using a backward AIC step function.  Further reductions were made by evaluating high VIFs versus predictor significance and making a judgement based its effect on adjusted $R^2$.

\[
Y = \beta_R x_R + \beta_{X3B} x_{X3B} + \beta_{CS} x_{CS} + \beta_{RA} x_{RA} + \beta_{CG} x_{CG} + \beta_{SHO} x_{SHO} + \beta_{SV} x_{SV} + \beta_{HRA} x_{HRA} + \beta_{salary} x_{salary}
\]

```{r}
teams_small = subset(teams, select = c(W, R, H, X2B, X3B, HR, CS, SF, RA, CG, SHO, SV, HA, HRA, BBA, E, FP, salary, GIDP, WHIP))
```

